Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-24 article/main.tex
fixed comments version 0.3

@@

The word \textit{startup} appeared in the SE literature for the first time in 
1994 in an article written by Carmel \cite{Camel1994a} where he studied the 
\textit{time-to-completion in a young package firm}. %MUN: If this is a quote 
%from Carmel, put it in quotation marks, like you did in the previous section 
%with Suttons quote. Generally, check that you %are consistent how you use 
%quotations throughout the paper.
He noticed how these companies were particularly innovative and successful, 
advocating for the need of more research investigating software development 
practices so as to replicate success and try to transfer it to other technology 
sectors.

@@

Additionally, the role of SPI has been neglected because seen as an obstacle to 
the development team's creativity and flexibility \cite{Coleman2008a}, and to 
the need of a quick delivering product process environment. %MUN: the second
%part of the sentence is just a statement without reference. If it refers to a 
%statement by Coleman, put his reference at the end OR find the correct 
%reference OR delete the statement.
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and Mirel 
\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in 
internet startups, are oriented to usability and scalability, even though the 
market and application type heavily influences the quality-demand 
\cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable 
functionality, studies suggest to externalize the complexity of parts of the 
project to third party solutions by means of outsourcing activities, software 
reuse and open-source strategies \cite{Hanna2010, Jansen2008, Wall2001, 
Bean2005}. %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source 
%strategies \cite{Bla}.

%------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new area of the SE research, trying to tackle the problem of 
the \textit{technical debt},  brings and encompasses different implications in 
studying development in software startups \cite{Tom2013}. The metaphoric 
neologism was originally introduced by Cunningham in 1992 
\cite{TechnicalDebtCunn} and has recently attracted the attention of SE 
researchers\footnote{To attest the fact that \textit{technical debt} is gaining 
traction among researchers, we mention two important contributions which 
characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} 
and a dedicated workshop \cite{workshopDebt} organized by the Software 
Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is 
well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea 
is that developers sometimes accept compromises in a system in one dimension 
(e.g., modularity) to meet an urgent demand in some other dimension (e.g., a 
deadline), and that such compromises incur a ``debt'' on which ``interest'' has 
to be paid and which the ``principal'' should be repaid at some point for the 
long-term health of the project''}.  The compromise between high-speed and 
high-quality engineering is faced daily by software startups, not only in terms 
of architecture design but in multifaceted aspects (weak project management, 
testing, process control).

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}
\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby 
%to keep the height of the figure, expanding however the width (use horizontal 
%space since it is a two column figure). Change also in figure:
% - Write "Data Analysis" horizontally
% - Change names of macro-phases to upper-case, i.e. Design and Execution, 
%   Data Collection, Data Analysis, etc.
% - Add section number to macro-phase name, i.e. 3.1 Design and Execution, 3.2 
%   Data Collection
% - Change "Discussion and Validation" into "Theory Validation" as you name it 
%   in the text.
%Also, add a section (3.5) that describes shortly what you did in "Theory 
%Validation"

@@

All companies, except C10, were founded within the last three years (2009-2012), 
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct?
by an average of 3 starting members, which were in majority developers. 
Moreover, the number of current employees shows how, to different degrees, 
companies expanded the initial teams. All companies, except C5, brought their 
first product to market in less than 6 months from the idea conception. The 
products consist of web-applications, launched in different nations and market 
sectors. The growing team size and the publicly available data suggest an 
healthy status of the businesses.

@@

The follow-up questionnaires were designed to capture additional data, gather 
missing information and confirm interview results by triangulation. The 
questionnaires have been tailored to each startup, partially taking advantage of 
the repertory grid principles \cite{Edwards2009}. To obtain better chances of 
quickly obtaining responses, questionnaires have been sent to respondents 
immediately after transcribing the interview results. Two weeks after the 
conclusion of the last interview, we closed our follow-up questionnaires and 
collected the data. %MUN: I don't understand this: you
%collected the data from questionnaires after ALL interviews have been 
%finished, but on the other hand you said the overall process was iterative 
%(data analysis and triangulation let themes emerge which in turn were used
%in subsequent interviews).

@@

\subsection{Theory Validation}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e. 
%the section following the GSM)

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical 
categories using the \textit{italic} font style. For the main categories shown 
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. 
CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up 
development}, which is the most interconnected node in the framework reflecting 
the fact that \textit{``it is the one (category) with the greatest explanatory 
power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that 
reach the core category (efficiency, effectiveness and performance) is discussed 
in detail in section \ref{valt}. %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?

@@

The \textit{severe lack of resources} forces the company to focus on 
implementing an essential set of functionalities and it is one of the main 
reasons why the \textit{product quality has low priority} respect to other more 
urgent needs\footnote{There are some exceptions where the quality aspects 
actually matter and such cases will be discussed in section \ref{valt}%MUN: 
%this refers to the validity threats section. Do you mean rather the "Theory 
%Validation" section?
.}. At the same time, to be able to deal with such constraints, startups must 
depend on a small group of capable and motivated individuals.

@@

A \textit{limited access to expertise} forces the team to rely mainly on their 
personal abilities, even though asking opinions to mentors is reported to be a 
viable practice to maintain feasible objectives. Furthermore \textit{teams work 
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to 
%this statement since it is a citation? 
mainly constrained by a tight \textit{time shortage}.

@@

To achieve a good level of UX while dealing with lack of human resources and 
time shortages, startups can analyze similar products of bigger companies which 
can afford more rigorous usability studies. Then, the users' feedback and 
product metrics begin to have a central role in determining the level of UX 
achieved. Product metrics are typically web-based \textit{statistical hypothesis 
testing}, such as A/B testing. %MUN: add reference or explain shortly what A/B 
%testing is

@@

\item The \textit{efficiency emerges after using the product}, letting 
engineers avoid wasting time in excessive improvements of not-validated 
functionalities. The level of efficiency can be optimized after attesting the 
effectiveness of the minimal set of functionalities, obtained according to CAT2 
and to the concept of MVP.
\item The \textit{product should be reasonably ready-to-scale} to be able to 
accommodate a potential growth of the user-base\footnote{Note that startups 
tackle fast growing markets which are particularly subject to sudden user 
growths}. Thanks to modern cloud services, startups can \textit{externalize 
complexity to third party solutions} achieving a good level of scalability with 
a reasonable effort.
\item Realizing high reliability is not an urgent priority since \textit{users 
are fault-tolerant in innovative beta products}. In these cases, users have 
typically a positive attitude towards the product, even though it presents 
unreliable behaviours. In this regard, the main focus of beta testing is 
reducing frictions between the product and the users, often incorporating 
usability testing. In fact, the beta release is typically the first time that 
the software is available outside of the organization that developed it. 
\footnote{Detailed discussion of the impact of innovative products on the user 
satisfaction is presented in Appendix according to the Kano model.}.
%MUN: do you use the Kano model in Theory Validation? Change text to refer to 
%the correct section. [After I've read the respective section, I'd suggest to 
%remove the Kano discussion altogether to shorten the paper. Keep the footnote, 
%reference the correct section, but remove "according to the Kano model]
\end{compactitem}

@@

Furthermore, a rigid project management is perceived as \textit{``waste of 
time''} which hinders the development speed since the \textit{uncertainty makes 
formal scheduling pointless} (C9  reports that \textit{``initial chaos helps to 
develop faster''}). Startups' \textit{minimal project management}  is supported 
by keeping: \textit{internal milestones short and informal}, 
\textit{low-precision task assignment mechanism}  and a very low attention for 
project metrics (paraphrasing C13, \textit{``the only track of progress was made 
by looking at closed tickets''}). In this context \textit{only a final release 
milestone is viable}, which helps practitioners to remain focused on short term 
goals and put new features in production. To further \textit{speed-up 
development}, startups \textit{simplify issue management by integrating it with 
feature management}. Then, the \textit{limited number of functionalities} and 
the {use of standard/known technologies} with a \textit{simple workflow} are the 
main reasons for 
not establishing a heavy project management process since the project and 
technologies can be autonomously managed by the development team (with 
characteristics of: co-location, self-organization and very small size). %MUN: 
%do you mean small team size?

@@

As discussed in section \ref{valt} %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?
, we formed the theory by applying the paradigm model, introduced by Corbin 
\cite{Corbin1990}. The theory is composed by different elements reflecting the 
paradigm model:

@@

In this section we discuss the validity of the GSM by means of 
cross-methodological observations:
%MUN: expand this discussion a bit (motivation why you do this like you do it) 
%and move to Section 3.5. Then write here just that you perform theory 
%validation as discussed in Section 3.5 (and remove the bullet list below).

\begin{compactitem}
\item Comparison between the GSM and similar  models and frameworks.
\item Comparison between the GSM and literature. We highlight 
the areas which have been neglected by existing studies, providing possible 
directions for future studies.
\item Discussion of confounding factors which could interfere with the GSM.
\item Validation of the correctness of the GT procedure by attesting relations 
between the GSM categories, evaluating if they are compliant with the empirical 
data we collected.

@@

To validate the generalization of the theory, we describe conceptualizations 
defined from the GSM that are in support of a previous model developed by 
Coleman \cite{Coleman2007,Coleman2008a, Coleman2008}. We refer to Coleman's work 
since only he has, to the best of our knowledge, conducted a similar research in 
startups' context, even though with a different focus. Coleman investigated 
factors in software development that hinder initiatives of software process 
improvement (SPI) in much more mature companies. %MUN: so he did not do research 
%in a startup context? You need to clarify what you mean by "more mature 
%companies"... I assume you still mean startups, but startups at a later stage, 
%i.e. after first product release.

@@

Coleman's framework aims to highlight how managers consider two distinct kinds 
of processes: \textit{essentials} and \textit{non-essentials}. The 
\textit{essential} processes are the most closely linked to product development, 
such as requirements gathering, design and testing. The \textit{non-essential} 
processes are those that might be omitted, such as planning, estimating and 
staging meetings. In particular, he discusses how practices are routinely 
removed: \textit{``With most methodologies and approaches, very few stick to the 
letter of them and they are always adapted, so we adapted ours to the way we 
wanted it to work for us, for our own size and scale''.} %MUN: add ref to 
%Coleman's paper where he states this

@@

As also reported in the GSM, the definition of a \textit{``minimum process''} 
is not a matter of poor knowledge and training, but rather it is the necessity 
of operating with solutions that let the company move faster. 
\textit{``One-size-fits-all''} solutions have always found difficulty in 
penetrating small software organizations \cite{Staples2007}. When startups began 
establishing any SPI process, they experienced \textit{process erosions} %MUN: 
%add ref (Coleman?). You should add, for each category you use from Coleman, 
%add the reference to the paper. This makes it clear that the category comes 
%from him (and is not one of our categories)
, which resolved to a barely sufficient workflow to satisfy the organizational 
business needs.

Then, software startups favor the use of agile principles in support of 
\textit{creativity} and \textit{flexibility} instead of SPI methods, whereby 
processes need to be predictable and repeatable. Nimble and ad-hoc solutions 
prevent the use of heavy bureaucracy and fast communication strategies %MUN: 
%seems to contradict each other: "ad-hoc solutions prevent the use of heavy 
%bureaucracy AND fast communication strategies" I agree that ad-hoc solutions 
%do not produce heavy bureaucracy, but they allow fast communication 
%strategies... I may be wrong, but then you need to explain it better
, even though the accumulated tacit knowledge is hard to manage and to transfer 
to new hires as also discussed in category \textit{CAT6}.

@@

In conclusion, by the comparison of the GSM with Coleman's framework, we have 
found that despite Coleman's study contributes to define which factors hinder 
the introduction of SPI %MUN: add: "of SPI in startups"?
, his results are coherent with the theory generated within our research %MUN: 
%add "our research, focused at explaining software development in general in 
%startups." ?
. In this regard, a wide variability of processes is a key factor for startups. 
In contrast with some SPI processes, such as \textit{Six Sigma} which objective 
is minimizing unpredictability in the definition of workflows \cite{Sixsigma}, 
startups seem to follow the \textit{Fischer's Fundamental Theorem of natural 
selection} \cite{Fisher}. Moving towards flexible and variable processes (easy 
to adapt to the uncertain conditions), increase the odds of ``natural 
selection'' only when some ``restrictive conditions'' are met. In other words, 
even though ignoring SPI practices is advisable %MUN: under which 
%circumstances? This statement sounds too general, i.e. don't do SPI ever...
%And the following part,I did not fully understand. Try to clarify.
not following SPI does not predict major adaptiveness of startups development 
to the market. But, to imply a major evolutionary product/market fit, certain 
conditions explained in GSM need to be considered.

%MUN: As I wrote already in the previous draft, you need to clarify at the 
%beginning of this subsection which frameworks you are going to compare GSM to. 
%Now, you say that Coleman is the only author who did research, comparable to 
%yours, into startups. Now there comes Baskerville and Brooks. Write at the 
%beginning of this section, shortly, which frameworks/authors you compare GSM 
%to and why, then create subsections for each. 
Also Baskerville \cite{Internet} refers to SPI as a development approach 
typically effective only in large-scale, long-term development efforts with 
stable and disciplined processes. By applying a grounded theory analysis in 10 
companies he found that major causal factors that influence development are:

@@

As discussed in CAT7, 1) can be mapped to \textit{time shortage} and 3) to 
\textit{limited access to expertise}, whilst  2) can refer to \textit{uncertain 
conditions make long-term planning not viable}, explained in CAT2. Even though 
with different research focus and context of study % - understand how and why 
%Internet-speed software development (i.e. an agile method oriented using daily 
%builds aimed at developing a product with high speed) differs from traditional 
%software development - MUN: put this sentence where you introduce 
%Baskerville's work
, Baskerville revealed similar causal factors as the GSM.

@@

\item The most urgent priority of software development is to shorten 
time-to-market. When developing a new product, which has not been attempted 
before, it is crucial for the company to release the product soon and observe 
the users reactions. The company's chances to survive, are strictly related to 
the ability to find the right product/market fit as quickly as possible. 
Developing software without releasing it for a long period of time, coming up 
with very complex solutions often is not a viable option. Business demands, 
investors interests, market pressures and the general uncertain conditions force 
startups to quickly iterate on the product by exploring and experimenting new 
functionalities progressively rolled-out to early-adopters.

>>>>

%CG: Not fixed. There is a huge gap on the topic, the refs in the introduction 
%are those retrieved by the SMS.

@@

The word \textit{startup} appeared in the SE literature for the first time in 
1994 in an article written by Carmel \cite{Camel1994a} where he studied the 
time-to-completion in a young package firm. %MUN: If this is a quote 
%from Carmel, put it in quotation marks, like you did in the previous section 
%with Suttons quote. Generally, check that you %are consistent how you use 
%quotations throughout the paper. 
%CG: fixed, btw  it was just to emphasize his the focus of his study. 
He noticed how these companies were particularly innovative and successful, 
advocating for the need of more research investigating software development 
practices so as to replicate success and try to transfer it to other technology 
sectors.

@@

Additionally, the role of SPI has been neglected because seen as an obstacle to 
the development team's creativity and flexibility, and to 
the need of a quick delivering product process environment \cite{Coleman2008a}.%MUN: the second part of the sentence is just a statement without reference. If it refers to a 
%statement by Coleman, put his reference at the end OR find the correct 
%reference OR delete the statement.
%CG fixed
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and Mirel 
\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in 
internet startups, are oriented to usability and scalability, even though the 
market and application type heavily influences the quality-demand 
\cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable 
functionality, studies suggest to externalize the complexity of parts of the 
project to third party solutions by means of outsourcing activities 
\cite{Hanna2010}, software reuse \cite{Jansen2008} and open-source strategies , \cite{Wall2001, Bean2005}. %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source 
%strategies \cite{Bla}.
%CG: fixed
%------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new area of the SE research, trying to tackle the problem of 
the \textit{technical debt},  brings and encompasses different implications in 
studying development in software startups \cite{Tom2013}. The metaphoric 
neologism was originally introduced by Cunningham in 1992 
\cite{TechnicalDebtCunn} and has recently attracted the attention of SE 
researchers\footnote{To attest the fact that \textit{technical debt} is gaining 
traction among researchers, we mention two important contributions which 
characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} 
and a dedicated workshop \cite{workshopDebt} organized by the Software 
Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is 
well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea 
is that developers sometimes accept compromises in a system in one dimension 
(e.g., modularity) to meet an urgent demand in some other dimension (e.g., a 
deadline), and that such compromises incur a ``debt'' on which ``interest'' has 
to be paid and which the ``principal'' should be repaid at some point for the 
long-term health of the project''}.  The compromise between high-speed and 
high-quality engineering is faced daily by software startups, not only in terms 
of architecture design but in multifaceted aspects (weak project management, 
testing, process control).

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=5in]{figures/completemethodology2}
\caption{Grounded theory study process overview}
\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby
%to keep the height of the figure, expanding however the width (use horizontal
%space since it is a two column figure). Change also in figure:  - Write "Data
%Analysis" horizontally  - Change names of macro-phases to upper-case, i.e.
%Design and Execution,     Data Collection, Data Analysis, etc.  - Add section
%number to macro-phase name, i.e. 3.1 Design and Execution, 3.2     Data
%Collection  - Change "Discussion and Validation" into "Theory Validation" as
%you name it     in the text. Also, add a section (3.5) that describes shortly
%what you did in "Theory  Validation" CG: fixed... fontsize cannot be increased
%uniformly among all the boxes, then I skipped that part...

@@

All companies, except C10, were founded within the last three years (2009-2012), 
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct? CG: right
by an average of 3 starting members, which were in majority developers. 
Moreover, the number of current employees shows how, to different degrees, 
companies expanded the initial teams. All companies, except C5, brought their 
first product to market in less than 6 months from the idea conception. The 
products consist of web-applications, launched in different nations and market 
sectors. The growing team size and the publicly available data suggest an 
healthy status of the businesses.

@@

The follow-up questionnaires were designed to capture additional data, gather 
missing information and confirm interview results by triangulation. The 
questionnaires have been tailored to each startup, partially taking advantage of 
the repertory grid principles \cite{Edwards2009}. To obtain better chances of 
quickly obtaining responses, questionnaires have been sent to respondents 
immediately after transcribing the interview results. Most of the follow-up questionnaires have been collected within one day after the 
conclusion of the last interview.   %MUN: I don't understand this: you
%collected the data from questionnaires after ALL interviews have been 
%finished, but on the other hand you said the overall process was iterative 
%(data analysis and triangulation let themes emerge which in turn were used
%in subsequent interviews). CG: fixed...

@@

%------------------------------------------------------------------------------%
\subsection{Theory Validation}
\label{rm:val}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e.
%the section following the GSM)
%CG: fixed

Presenting GT is challenging for a researcher, who must pay attention to structure the level of details included, and to the way data is portrayed to display evidence of the emergent categories.  In order to assess our research and determine if the GT is sufficiently grounded, we used a systematic technique to assess the theory, following evaluation lists and factors presented in  \cite{Strauss1998}.

First we had to consider the following factors:

\par\vspace {0.4cm}
\begin{compactitem}
\item \textit{Fit}, that describes the consistence of the theory with the real data.
\item \textit{Understanding}, that describes the degree to which participants agree with the theory.
\item \textit{Generality}, that describes the level of abstraction of the theory, and how sufficiently it can be adopted by practitioners without losing its relevance.
\item \textit{Control}, that describes how well the theory enables practitioners to fully understand the described situation.
\end{compactitem}
\par\vspace {0.4cm}

In order to achieve a high-level of fit between the theory and the raw empirical data, we started GT without any assumption on how the emergent patterns would influence the research direction. Ideas and perceptions evolved during the mapping study and interviews. In addition, constant comparative methods, overturning of some early categories as new data came to light, generation and testing of interim hypotheses, and constant re-evaluation of the interview transcripts, ensured that researchers' bias was minimized and theoretical fit maintained.

To represent understandable emergent theories, concepts and categories were carefully developed using appropriate and self-explanatory codes. The developing theory was not immediately presented to the participants to prevent potential bias in the responses. But rather, the theoretical framework has been presented and described to startups practitioners not involved in the data collection, whose reaction was positively expressed.

From generality perspective, we developed a theory from abstract categories, which allowed us to create a general guide to constantly changing situations. Nevertheless, we took attention on maintaining a reasonable level of abstraction, keeping focus on their sensitizing characteristics and providing low-level subgroup of categories.

Finally, we provided a comprehensive set of categories with detailed interrelations to explain how the \textit{core-category} is achieved and the reasons of its consequences, in order to enable practitioners to independently understand and analyze situations, predict changes and their consequences, and to be capable of revising actions or the theory itself. Using both methodological tools and features provided by \textit{AtlasTI}, we explored and tested:

\par\vspace {0.4cm}
\begin{compactitem}
\item Each category and the strength of relations between them.
\item Hypotheses, derived from and related to the emergent theory.
\item Deviant cases to ensure robustness and general applicability.
\end{compactitem}
\par\vspace {0.4cm}

Additionally, Strauss and Corbin provided a list of questions to assist in determining how well the findings are grounded \cite{Strauss1998}. They questions are as follows:

\par\vspace {0.4cm}
\begin{compactenum}
\item Are concepts generated, and are the concepts systematically related?
\item Are there many conceptual linkages and are the categories well developed?
\item Is variation\footnote{Variation refers to the variety of contexts which theory can be applied.} built into the theory and are the conditions under which variation can be found built into the study and explained?
\item Are the conditions under which variation can be found built into the study and explained?
\item Has the process been taken into account, and does the theory stand the test of time?
\item Do the theoretical findings seem significant, and to what extent?
\end{compactenum}
\par\vspace {0.4cm}

Answering the above questions, the concepts were generated according to the coding process. They were systematically related through the use of a network diagram (see figure \ref{fig:gt:completemethodology}), by which we established the linkages and relations between the concepts. At each iteration of the grounded theory process, we considered and examined a concept within different conditions and dimensions, trying to incorporate opinions of a broader as possible range of practitioners.

All the linkages and categories were constructed by the use of Atlas.TI and compared according to the data analysis process. Variations of the generated theory were applied according to raw grounded data with the aid of codes and memos. Moreover during this study extensive explanation of the concepts were presented in-vivo statements as reported by practitioners.

The process has been designed by the researches in different steps, explaining and making clear the purpose of each of them. Thus, the same processes enable other research to conduct the same study within similar contexts. Moreover \textit{comparison with the state-of-the-art} has been applied in order to validate the theory and enhance odds of its applicability within a wider time-frame. By this comparison we  highlight the areas which have been neglected by existing studies, providing possible directions for future studies (see subsection \ref{sect:theory:validation:others}). Moreover we study the confounding factors which could interfere with the application of the GSM model (see subsection \ref{sect:an:comp:cat-vs-literature:conf}).

Then, we performed an \textit{evaluation with empirical data} (see subsection \ref{sect:theory:validation:stats}), quantifying scores for each company involved in our empirical study by defining metrics based on a set of rubrics and evaluating startups accordingly, through an analysis of interview transcripts and codes. The complete statistical procedure and provided  rubrics  to quantify the measures are illustrated in Appendix.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical 
categories using the \textit{italic} font style. For the main categories shown 
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. 
CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up 
development}, which is the most interconnected node in the framework reflecting 
the fact that \textit{``it is the one (category) with the greatest explanatory 
power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that 
reach the core category (efficiency, effectiveness and performance) is discussed 
in detail in section \ref{res:val}. %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?
%CG:fixed

@@

The \textit{severe lack of resources} forces the company to focus on 
implementing an essential set of functionalities and it is one of the main 
reasons why the \textit{product quality has low priority} respect to other more 
urgent needs\footnote{There are some exceptions where the quality aspects 
actually matter and such cases will be discussed in section \ref{res:val}%MUN: 
%this refers to the validity threats section. Do you mean rather the "Theory 
%Validation" section?
%CG fixed
.}. At the same time, to be able to deal with such constraints, startups must 
depend on a small group of capable and motivated individuals.

@@

A \textit{limited access to expertise} forces the team to rely mainly on their 
personal abilities, even though asking opinions to mentors is reported to be a 
viable practice to maintain feasible objectives. Furthermore \textit{teams work 
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to 
%this statement since it is a citation? 
%CG: not fixed. It represents a subcategory, expressed by the 70% of the interviewed startups
mainly constrained by a tight \textit{time shortage}.

@@

To achieve a good level of UX while dealing with lack of human resources and 
time shortages, startups can analyze similar products of bigger companies which 
can afford more rigorous usability studies. Then, the users' feedback and 
product metrics begin to have a central role in determining the level of UX 
achieved. Product metrics are typically web-based \textit{statistical hypothesis 
testing}, such as A/B testing \cite{AB}. %MUN: add reference or explain shortly what A/B 
%testing is
%CG: fixed

@@

\item The \textit{efficiency emerges after using the product}, letting 
engineers avoid wasting time in excessive improvements of not-validated 
functionalities. The level of efficiency can be optimized after attesting the 
effectiveness of the minimal set of functionalities, obtained according to CAT2 
and to the concept of MVP.
\item The \textit{product should be reasonably ready-to-scale} to be able to 
accommodate a potential growth of the user-base\footnote{Note that startups 
tackle fast growing markets which are particularly subject to sudden user 
growths}. Thanks to modern cloud services, startups can \textit{externalize 
complexity to third party solutions} achieving a good level of scalability with 
a reasonable effort.
\item Realizing high reliability is not an urgent priority since \textit{users 
are fault-tolerant in innovative beta products}. In these cases, users have 
typically a positive attitude towards the product, even though it presents 
unreliable behaviours. In this regard, the main focus of beta testing is 
reducing frictions between the product and the users, often incorporating 
usability testing. In fact, the beta release is typically the first time that 
the software is available outside of the organization that developed it. 
\footnote{Detailed discussion of the impact of innovative products on the user 
satisfaction is presented in subsection \ref{sect:an:comp:cat-vs-literature:conf}}.
%MUN: do you use the Kano model in Theory Validation? Change text to refer to 
%the correct section. [After I've read the respective section, I'd suggest to 
%remove the Kano discussion altogether to shorten the paper. Keep the footnote, 
%reference the correct section, but remove "according to the Kano model]
%CG: fixed
\end{compactitem}

@@

Furthermore, a rigid project management is perceived as \textit{``waste of 
time''} which hinders the development speed since the \textit{uncertainty makes 
formal scheduling pointless} (C9  reports that \textit{``initial chaos helps to 
develop faster''}). Startups' \textit{minimal project management}  is supported 
by keeping: \textit{internal milestones short and informal}, 
\textit{low-precision task assignment mechanism}  and a very low attention for 
project metrics (paraphrasing C13, \textit{``the only track of progress was made 
by looking at closed tickets''}). In this context \textit{only a final release 
milestone is viable}, which helps practitioners to remain focused on short term 
goals and put new features in production. To further \textit{speed-up 
development}, startups \textit{simplify issue management by integrating it with 
feature management}. Then, the \textit{limited number of functionalities} and 
the {use of standard/known technologies} with a \textit{simple workflow} are the 
main reasons for 
not establishing a heavy project management process since the project and 
technologies can be autonomously managed by the development team (with 
characteristics of: co-location, self-organization and very small team size). %MUN: 
%do you mean small team size?
%CG: fixed

@@

As discussed in section \ref{res:val} %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?
%CG:fixed
, we formed the theory by applying the paradigm model, introduced by Corbin 
\cite{Corbin1990}. The theory is composed by different elements reflecting the 
paradigm model:

@@

In this section we discuss the validity of the GSM by means of 
cross-methodological observations, as discussed in section \ref{rm:val}.
%MUN: expand this discussion a bit (motivation why you do this like you do it) 
%and move to Section 3.5. Then write here just that you perform theory 
%validation as discussed in Section 3.5 (and remove the bullet list below).
%CG:fixed

@@

To validate the generalization of the theory, we describe conceptualizations 
defined from the GSM that are in support of previous models developed by 
Coleman \cite{Coleman2007,Coleman2008a, Coleman2008}, Baskerville \cite{Internet} and finally Brooks \cite{BrooksJr1987}.

Starting from the Coleman's work, we refer to him since only he has, to the best of our knowledge, 
conducted a similar research in startups' context, even though with a 
different focus. Coleman investigated 
factors in software development that hinder initiatives of software process 
improvement (SPI) in a later stage, representing also companies in expansion phase with more than 100 employees. %MUN: so he did not do research 
%in a startup context? You need to clarify what you mean by "more mature 
%companies"... I assume you still mean startups, but startups at a later stage, 
%i.e. after first product release.
%CG: fixed

@@

Coleman's framework aims to highlight how managers consider two distinct kinds 
of processes: \textit{essentials} and \textit{non-essentials}. The 
\textit{essential} processes are the most closely linked to product development, 
such as requirements gathering, design and testing. The \textit{non-essential} 
processes are those that might be omitted, such as planning, estimating and 
staging meetings. In particular, he discusses how practices are routinely 
removed: \textit{``With most methodologies and approaches, very few stick to the 
letter of them and they are always adapted, so we adapted ours to the way we 
wanted it to work for us, for our own size and scale''} \cite{Coleman2008}. %MUN: add ref to 
%Coleman's paper where he states this
%CG fixed

@@

As also reported in the GSM, the definition of a \textit{``minimum process''} 
is not a matter of poor knowledge and training, but rather it is the necessity 
of operating with solutions that let the company move faster. 
\textit{``One-size-fits-all''} solutions have always found difficulty in 
penetrating small software organizations \cite{Staples2007}. When startups began 
establishing any SPI process, they experienced \textit{process erosions} \cite{Coleman2008} %MUN: 
%add ref (Coleman?). You should add, for each category you use from Coleman, 
%add the reference to the paper. This makes it clear that the category comes 
%from him (and is not one of our categories)
%CG: fixed
, which resolved to a barely sufficient workflow to satisfy the organizational 
business needs.

Then, software startups favor the use of agile principles in support of 
\textit{creativity} and \textit{flexibility} instead of SPI methods, whereby 
processes need to be predictable and repeatable. Nimble and ad-hoc solutions 
prevent the use of heavy bureaucracy and formal communication strategies %MUN: 
%seems to contradict each other: "ad-hoc solutions prevent the use of heavy 
%bureaucracy AND fast communication strategies" I agree that ad-hoc solutions 
%do not produce heavy bureaucracy, but they allow fast communication 
%strategies... I may be wrong, but then you need to explain it better
%CG:fixed, you are right :) ...  
, even though the accumulated tacit knowledge is hard to manage and to transfer 
to new hires as also discussed in category \textit{CAT6}.

@@

In conclusion, by the comparison of the GSM with Coleman's framework, we have
found that despite Coleman's study contributes to define which factors hinder
the introduction of SPI in startups %MUN: add: "of SPI in startups"? CG: fixed ,
his results are coherent with the theory generated within our research, focusing
at explaining the overall strategies within software development in
startups%MUN:  %add "our research, focused at explaining software development in
general in  %startups." ? CG:fixed . In this regard, a wide variability of
processes is a key factor for startups.  In contrast with some SPI processes,
such as \textit{Six Sigma} which objective  is minimizing unpredictability in
the definition of workflows \cite{Sixsigma},  startups seem to follow the
\textit{Fischer's Fundamental Theorem of natural  selection} \cite{Fisher}.
Moving towards flexible and variable processes (easy  to adapt to the uncertain
conditions), increase the odds of ``natural  selection'' only when some
``restrictive conditions'' are met. In other words,  even though ignoring SPI
practices is advisable, under circumstances of high uncertainty and severe lack
of resources, %MUN: under which  %circumstances? This statement sounds too
general, i.e. don't do SPI ever... %CG: fixed %And the following part,I did not
fully understand. Try to clarify. CG:fixed  not following SPI does not predict
major adaptiveness of startups development  towards the customer's needs.
Nevertheless, to imply a major evolutionary product/market fit, certain
conditions explained in GSM need to be considered.

%MUN: As I wrote already in the previous draft, you need to clarify at the 
%beginning of this subsection which frameworks you are going to compare GSM to. 
%Now, you say that Coleman is the only author who did research, comparable to 
%yours, into startups. Now there comes Baskerville and Brooks. Write at the 
%beginning of this section, shortly, which frameworks/authors you compare GSM 
%to and why, then create subsections for each.

%CG: partially fixed. Agreed on the beginning of the section, not agreed on the division in subsections. Let's discuss it by a skype call.

Also Baskerville \cite{Internet} refers to SPI as a development approach
typically effective only in large-scale, long-term development efforts with
stable and disciplined processes.Internet-speed software development (i.e. an
agile method oriented using daily builds aimed at developing a product with high
speed) differs from traditional software development. Baskerville, by applying a
grounded theory analysis in 10 companies, found that major causal factors that
influence development are:

@@

As discussed in CAT7, 1) can be mapped to \textit{time shortage} and 3) to 
\textit{limited access to expertise}, whilst  2) can refer to \textit{uncertain 
conditions make long-term planning not viable}, explained in CAT2. Even though 
with different research focus and context of study % - understand how and why 
%Internet-speed software development (i.e. an agile method oriented using daily 
%builds aimed at developing a product with high speed) differs from traditional 
%software development - MUN: put this sentence where you introduce 
%Baskerville's work. %CG: fixed
, Baskerville revealed similar causal factors as the GSM.

@@

\item External validity - which requires establishing the domain to which the study's findings can be generalized (partially discussed above).
